\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-09-20 Manuscript/manuscript.tex
rewrote abstract and intro

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun concluded that the reaction times  
from the 2022 World Championships seemed to be generally faster compared to the 
other reaction times they considered, but they did not perform any formal 
statistical analysis. This paper questions if athletes who competed at the 2022
World Track and Field Championship had similar times at other competitions and
if 0.1 seconds is a reasonable disqualification barrier. To do so, we employ a 
signed-rank test for clustered data to compare reaction times for the same 
athletes at different competitions and a generalized linear mixed model 
with random venue (year) effect and a random heat effect within venue 
in order to model the reaction time data. This matter needs to be addressed 
because disqualification based on allowable reaction time will continue to be an
issue in future world championships.

\bigskip\noindent{\sc Keywords}:
False Start, Hurdles, Reaction Time, Seiko

@@

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle. After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  The
reaction time barrier for all sprint related events is 0.1 seconds as set by the
International Association of Athletics Federation (IAAF) and is meant
to prevent athletes from trying to predict the starting gun.
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold, was met
with shock from both Allen and fans alike. NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they realized that Allen was disqualified, angry at the result. Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete.

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships. However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website. He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed descriptive statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

The first objective of this paper is to determine whether there were 
abnormalities with athlete's reaction times at the 2022 World Track and Field
Championship.  Our analysis consists of quantitatively comparing reaction times for athletes who 
have competed at additional competitions to the 2022 World Track and Field 
Championship (Similar to a matched pairs design). By comparing reaction times from the same athlete across multiple
competitions we are able to determine if the 2022 reaction times differed 
significantly from those of other competitions.  This investigation, while
motivated by Allen's disqualifcation has wider impacts across track and field
that question if there is more variation than previously thought in track and
field timing equipment.

The second objective of this paper is to analyze whether the current
reaction time barrier of 0.1 seconds is fair, and in this regard there have been
many publications debating the human capacity for reaction times 
\of{This sounds funny but I'm not sure how to fix it}.
\citet{pain2007sprint} examined reaction times for nine male athletes. The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds. Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
These papers support increasing the reaction time barrier as they determined
that athletes are better than how IAAF thinks they are. Unfortunately, IAAF has
not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
Additionally, \citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been shown to be a reliable
measure of reaction time.

However, not all studies have been in favor of a decrease in the
reaction time barrier \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article. A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}. However it is
worth noting that \citet*{brosnan2017effects} and \citep{lipps2011implications}
were retroactive studies of athletes at events while \citet{pain2007sprint} and
\citet{komi2009iaaf} were experiments showing that there are various methods
of testing reaction time capability.

The second part of this paper will model the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects: 
one effect for the year and a second effect for the heat nested within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1).

The rest of this paper is organized as follows. Section~\ref{sec:Data} details 
how the two types of data were collected, as the signed-rank test as described
in Section~\ref{sec:rank} uses slightly different data than the generalized 
linear mixed-effect model (GLMM) that is further developed in 
Section~\ref{sec:glmm}. Section~\ref{sec:Results} provides the results of the two
analyses and ends by questioning the reaction time barrier. Lastly, 
Section~\ref{sec:concludingremarks} discusses the impact and limitations of the
paper. All the data and code of our analysis are available in a supplement.

>>>>

% Abstract should be under 200 words

\begin{abstract}
Devon Allen was disqualified from the men's 110-meter hurdle final at
the 2022 World Track and Field Championships after recording a
reaction time of 0.099 seconds—just 0.001 seconds faster than the
allowable threshold of 0.1 seconds. This paper explores whether the
reaction times of athletes at the 2022 Championships were unusually
fast compared to other competitions and evaluates the appropriateness
of the 0.1-second disqualification barrier. To investigate these
issues, we employ a signed-rank test for clustered data to compare
athletes' reaction times across multiple competitions, along with a
generalized linear mixed model (GLMM) that accounts for both venue and
heat effects. The results indicate that athletes at the 2022
Championships had significantly faster reaction times on average. This
analysis raises important questions about the validity of the current
reaction time threshold and suggests potential modifications to the
rules that could prevent future disqualifications under similar
circumstances.

\bigskip\noindent{\sc Keywords}:
False Start, GLMM, Hurdle, Reaction Time, Seiko, rank-based test

@@

Devon Allen, a University of Oregon alumnus, was a strong 
contender for the 2022 World Track and Field Championships 
in the 110-meter hurdle event. After running the event in 
12.84 seconds in June 2022, just 0.04 seconds short of the 
world record \citep{wa2022preview}, and placing third at 
the U.S. Track and Field Championships, Allen was well-prepared 
for the World Championships in August, also held in Eugene, 
Oregon. He advanced through the preliminary heats and semifinals 
with reaction times of 0.123 and 0.101 seconds, respectively. 
However, in the final heat, Allen was disqualified for a 
reaction time of 0.099 seconds, just 0.001 seconds faster 
than the 0.1-second threshold set by the International 
Association of Athletics Federations (IAAF). This decision 
was met with significant public outcry \citep{nbc2022devon}.

Allen's disqualification has been discussed extensively in online 
communities such as \url{www.LetsRun.com}, a website that functions 
as both a message board and a news outlet. The message board, 
similar to Reddit, is very active during major running events, 
including the Olympics and the World Track and Field Championships. 
The news section of the website featured articles such as 
\citep{johnson2022data} and \citep{johnson2022was} by Robert Johnson, 
the site’s founder. Johnson argued that there may have been an 
error with the timing equipment at the 2022 Championships. He 
cited graphs, descriptive statistics like median reaction times, 
and compared the reaction times at the 2022 Championships to 
those from other competitions, including the U.S. Championships 
and previous World Championships. However, these comparisons 
lacked formal statistical analysis. Johnson estimated a 99.95\% 
chance that Allen was wrongfully disqualified, based on his claim 
that the probability of the fastest reaction times from the men's 
100 meter, men's 110 meter hurdle, and women's 100 meter all 
occurring in the same year was approximately $(1/13)^3$ 
\citep{johnson2022was}.

The first objective of this paper is to determine whether athletes’ 
reaction times at the 2022 World Track and Field Championships 
were significantly different from other competitions. Our analysis 
compares reaction times for athletes who competed in multiple events, 
using a matched-pairs design. By comparing the same athletes' reaction
times across different competitions, we can assess whether the 2022 
reaction times differed significantly. While this investigation was 
motivated by Allen's disqualification, it also addresses broader concerns 
about potential variation in timing equipment across different venues
in short-distance dash or hurdle.

Given the observed variation in reaction times across venues, the 
question arises: is the 0.1-second disqualification threshold truly 
justified? Studies have presented evidence supporting both sides of 
the debate. \citet{pain2007sprint} examined reaction times in nine 
male athletes and found that the fastest athlete had an average 
reaction time of 0.087 seconds, with a standard deviation of 0.004 
seconds. Similarly, \citet{komi2009iaaf}, in a study commissioned by 
the IAAF, recommended lowering the disqualification threshold to 0.08 
or 0.085 seconds, arguing that athletes are capable of faster reaction 
times than the current threshold allows. On the other hand, some 
researchers have argued for increasing the reaction time threshold. 
\citet{brosnan2017effects} suggested raising it to 0.115 seconds based 
on data from European and World Championships, while a study of the 
2008 Olympics supported the conclusion that athletes cannot reliably 
react within 0.1 seconds \citep{lipps2011implications}. Despite these 
findings, the IAAF has not revised the threshold since its implementation 
in 1999. It is important to note that \citet{brosnan2017effects} and 
\citet{lipps2011implications} relied on retrospective event data, 
whereas \citet{pain2007sprint} and \citet{komi2009iaaf} conducted 
controlled experiments to assess reaction time capabilities.

The second objective of this paper is to model reaction times for 
non-disqualified athletes from the 1999 World Championships onward. 
We employ a gamma linear mixed-effects model with two random effects: 
one for the venue (year) and another for the heat nested within each venue.
This model allows us to estimate the probability of an athlete's 
reaction time falling below a specific threshold, such as 0.1 seconds. 
By examining reaction times across years and heats, we aim to assess 
the consistency of reaction times and whether the current 
disqualification threshold is statistically justified. This modeling 
approach provides a more detailed understanding of the factors that 
influence reaction times, offering insights into whether the 0.1-second 
threshold should be reconsidered for future competitions.

The rest of this paper is organized as follows. Section~\ref{sec:Data} 
describes how the data were collected. The signed-rank test, detailed 
in Section~\ref{sec:rank}, uses slightly different data from the 
generalized linear mixed-effects model (GLMM) developed in 
Section~\ref{sec:glmm}. Section~\ref{sec:Results} presents the results 
of both analyses and discusses the implications of the reaction time 
barrier. Finally, Section~\ref{sec:concludingremarks} outlines the 
paper’s impact and limitations. All data and code for our analysis 
are provided in the supplementary materials.
